EDA of Prosper Marketplace loan data by Zhaowei Liang
========================================================


# 1. Introduction and data wrangling

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(anytime)
library(dplyr)
library(GGally)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
raw <- read.csv("prosperLoanData.csv")
```

```{r}
#review variables
names(raw)
```

The dataset contains more than 100,000 loans't data with 81 features for each 
loan from Prosper Marketplace Loan company. The [dictionary](https://docs.google.com/spreadsheets/d/1gDyi_L4UvIrLTEC6Wri5nbaMmkGmLQBk-Yx3z0XDEtI/edit#gid=0) described what does each variable represents. The variables that worth 
investigating was picked by my limited knowledge in finance. However, 
later in the analysis, more variables might be added in analysis according 
to correlations between variables. 

```{r}
#pick features of interest can organize them in to a list
selected_features <- c('ListingKey', 'ListingCreationDate', 'Term', 'LoanStatus', 
                       'CreditGrade', 
                      'ListingCategory..numeric.', 'ProsperScore', 'BorrowerState',
                      'Occupation',
                      'EmploymentStatusDuration', 'IsBorrowerHomeowner', 'CreditScoreRangeLower', 'CreditScoreRangeUpper',
                      'IncomeRange', 'IncomeVerifiable', 'TotalCreditLinespast7years', 'TotalProsperLoans',
                      'LoanOriginationDate', 'LoanOriginalAmount', 'PercentFunded', 'Investors', 'StatedMonthlyIncome'
                     )
```

```{r}
#subset picked features from the raw dataset according to the list created in last chunk
loan_df = raw[, selected_features]
```

```{r}
str(loan_df)
```

```{r}
names(loan_df)
```

20 variables and 113937 observation are picked for analyzing. 

```{r}
#create a new column contains only years of all listing
loan_df$ListingCreationDate <- anydate(loan_df$ListingCreationDate)
loan_df$ListingCreationYear <- strftime(loan_df$ListingCreationDate, "%Y")
```

```{r}
loan_year <- table(loan_df$ListingCreationYear)
print(loan_year)
```

```{r}
#ggplot(data = loan_df, aes(ListingCreationYear)) + geom_bar(aes(x = ListingCreationYear, y = freq), stat="identity")
```

# 2. Univariate Plots Section

## 2.1 Borrowers' states
Geographical information is critical for the growth of a company. Firstly, analyze what state owns the most of loans on Proser Marketplace. However, since the population density is vary from state to state, this result may be less informative than people normally think.

```{r}
ggplot(loan_df,aes(x=reorder(BorrowerState,BorrowerState,
                     function(x)+length(x)))) + 
        geom_bar() + coord_flip() + theme(axis.text.y = element_text(size=5,vjust=1)) 
```

```{r}
table(loan_df$BorrowerState)
```

From the bar plot above, we can see that California has most total loans. 
Meanwhile, the total number of loans distributed similarly as population. 
One interesting find is there are 5515 loans has unkonwn state, are they 
from out of United States or it is data entry error? It is hard to allocate 
these loans to a specific states without additional data from Prosper Marketplace.

## 2.2 Occupation
Will loaner's occupation affects there chance of getting investors? 
Is there any occupation discriminate on Prosper MarketPlace? To better
answer these question, total amount of loans for different occupations 
are analyzed first. 

```{r}
ggplot(loan_df,aes(x=reorder(Occupation,Occupation,
                     function(x)+length(x)))) + 
        geom_bar() + coord_flip() + theme(axis.text.y = element_text(size=5,vjust=1)) 
```

Obviously, the dropdown menu on Prosper Marketplaces's website is not
exhaustive. The occupation "Other" siginificantly skewed the distribution.

## 2.3 Employment Status Duration

```{r}
sample(loan_df$EmploymentStatusDuration, 15)
```

```{r}
ggplot(data = subset(loan_df, !is.na(EmploymentStatusDuration)), aes(EmploymentStatusDuration)) + 
       geom_histogram(stat = "bin", binwidth = 10)
```

As shown in the graph above, most of the loaners provided employement 
data on this website do have a fairly short employment status history.

## 2.4 Credit Score
The credit score records is dvided into two features: 
1. CreditScoreRangeLower
2. CreditScoreRangeUpper
Firstly, the relationship between this two features is determined.

```{r}
loan_df$creditrange <- loan_df$CreditScoreRangeUpper - loan_df$CreditScoreRangeLower
table(loan_df$creditrange)
```

```{r}
table(loan_df$CreditScoreRangeUpper)
```

As show above, the the creditscorerangeupper is mostly 19 more than 
the creditscorerangelower. According to the properties of expected values.
Only one of these variables needs to analyzed. 

```{r}
ggplot(data = subset(loan_df, !is.na(CreditScoreRangeUpper)), aes(CreditScoreRangeUpper)) + 
       geom_histogram(stat = "bin", binwidth = 10)
```

As shown above, the distribution of credit score shows approximate normal 
distribution if we neglect outliers at 19.

## 2.5 Income Range

```{r}
table(loan_df$IncomeRange)
```

```{r}
ggplot(loan_df,aes(IncomeRange)) + 
        geom_bar() + coord_flip()
```

As show in the graph above, the income range of lenders is a good representation
of financial state of lenders. Quite a number of incomes is not displayed. But 
mostly the income range of lenders is concentrated below $50,000.

## 2.6 Total Credit Line

```{r}
ggplot(data = subset(loan_df, !is.na(TotalCreditLinespast7years)), aes(TotalCreditLinespast7years)) + 
       geom_histogram(stat = "bin", binwidth = 3)
```

## 2.7 Loan Origination Date
The purpose of analyzing loan origination date is verify of the data

```{r}
loan_df$LoanOriginationDate <- anydate(loan_df$LoanOriginationDate)
loan_df$LoanOriginationYear <- strftime(loan_df$LoanOriginationDate, "%Y")
```

```{r}
ggplot(loan_df,aes(LoanOriginationYear )) + 
        geom_bar()
```

## 2.8 Loan amount

```{r}
ggplot(data = subset(loan_df, !is.na(LoanOriginalAmount)), aes(LoanOriginalAmount)) + 
       geom_histogram(stat = "bin", binwidth = 4000)
```

```{r}
amount_list <- sort(table(loan_df$LoanOriginalAmount), decreasing = T)
head(amount_list, 20)
```

Most of the loan amount is on numbers that looks more "integer" like 4000, 15000 
insteand 15232 or 4355. Because the Prosper MarketPlace doesn't gurantee this 
amount will be fully funded, its make sense that most lenders put a approximate
number to take the most from investers.

## 2.10 Number of investors

```{r}
ggplot(data = subset(loan_df, !is.na(Investors)), aes(Investors)) + 
       geom_histogram(stat = "bin", binwidth = 60)
```

## 2.11 Loan status (Main Feature)
There are multiple status of loans per data dictionary:

Cancelled
Chargedoff
Completed
Current
Defaulted
FinalPaymentInProgress
Past due(Followed by buckets)

```{r}
levels(loan_df$LoanStatus)
```

```{r}
ggplot(loan_df,aes(LoanStatus)) + 
        geom_bar() + coord_flip()
```

As seen in this graph, the mojority of loans is in their "Current" Status. 
Becuase the loan status is also the main feature of our analysis, we need to 
go one step further on.

The loans status feature has 12 levels:

```{r}
levels(loan_df$LoanStatus)
```

To make the analsis easier in later stage, these levels will be divided into 
2 groups: Good loans and Bad loans, in the other word, the deliquency state of 
loans. After read the wikipedia and investopedia page of these words, they are 
divided as follow:

- Good loans():
    - Cancelled
    - Completed
    - Current
    - FinalPaymentInProgress
    - Past Due (1-15 days) 
    - Past Due (16-30 days)

- bad (deliquented) loans:
    - Chargedoff
    - Defaulted
    - Past Due (>120 days) 
    - Past Due (31-60 days) 
    - Past Due (61-90 days) 
    - Past Due (91-120 days)

A new column called "DeliquencyState" is created based on the creteria above.
Under this column, 1 represent loans with deliquencies and 0 represents loans
without deliquencies.

```{r}
loan_df$DeliquencyState <- ifelse(
                                      loan_df$LoanStatus == "Chargedoff" |
                                      loan_df$LoanStatus == "Defaulted" |
                                      loan_df$LoanStatus == "Past Due (>120 days)" |
                                      loan_df$LoanStatus == "Past Due (31-60 days)" |
                                      loan_df$LoanStatus == "Past Due (61-90 days)" |
                                      loan_df$LoanStatus == "Past Due (91-120 days)",
                                       'delinquent', 'GoodStanding')
```

```{r}
table(loan_df$DeliquencyState)
```

```{r}
ggplot(data = loan_df, aes(DeliquencyState, fill = factor(LoanStatus))) + geom_bar()
```

## 2.12 Univariate Analysis


### What is the structure of your dataset?

At the beginning of univariate analysis

### What is/are the main feature(s) of interest in your dataset?

The main features of interest in my analysis are the status of the loan and 
the number of investors. 

There are multiple status of loans per data dictionary:

Cancelled
Chargedoff
Completed
Current
Defaulted
FinalPaymentInProgress
Past due(Followed by buckets) ...

They are divided and regrouped into a new feature: DeliquencyState which is now the feature of interest.



### What other features in the dataset do you think will help support your \ investigation into your feature(s) of interest?

'ListingKey' 'ListingCreationDate' 'Term' 'LoanStatus' 'CreditGrade' 
'ListingCategory..numeric.' 'ProsperScore' 'BorrowerState' 'Occupation' 
'EmploymentStatusDuration' 'IsBorrowerHomeowner' 'CreditScoreRangeLower' 
'CreditScoreRangeUpper' 'IncomeRange' 'IncomeVerifiable' 
'TotalCreditLinespast7years' 'TotalProsperLoans' 'LoanOriginationDate' 
'LoanOriginalAmount' 'PercentFunded' 'Investors' 'ListingCreationYear' 
'creditrange' 'LoanOriginationYear' 'DeliquencyState'

### Did you create any new variables from existing variables in the dataset?

Yes, I created two variables to help my analysis:
1. LoanOriginationYear: this variable is parsed from LoanOriginationDate data in order to determine if the sample are radomly picked through years.

2. DeliquencyState: a binary column to determine the status of a loan. This variable is regrouped from loan status feature. 

### Of the features you investigated, were there any unusual distributions? \ Did you perform any operations on the data to tidy, adjust, or change the form \ of the data? If so, why did you do this?

I changed the data type of LoanOriginationDate to datetime so I can parse 
the year of loan origination much easier.




# 3. Bivariate Plots Section

This section is divided into 3 parts:
1. Relationship between loan status and support features.
2. Relationship between number of investors and supporting features
3. Relationship between some supporting features

## 3.1 Loan status (Deliquency State )

## 3.1.1 Deliquency State and  Employment Status Duration

```{r}
ggplot(data = loan_df, aes(x = DeliquencyState, y = CreditScoreRangeUpper)) + 
      geom_boxplot()
```

## 3.1.2 Deliquency State and  Income Range

```{r}
ggplot(data = loan_df, aes(IncomeRange, fill = factor(DeliquencyState))) + geom_bar() + coord_flip()
```

Although it is obvious that borrowers that has income range from 25,000 to 
49,999 has most of the delinquent loans, they also has most of loans in general. 
A more reasonable feature in this kind case would be the percent of deliquent 
loans compare to the total loans in that income range.

## 3.1.3 Deliquency State and  loan original amount

```{r}
ggplot(data = loan_df, aes(x = DeliquencyState, y = LoanOriginalAmount)) + 
      geom_boxplot()
```

In contraray to the intuition, delinquent loans has lower loan amount compare 
to good standing loans. I was expecting delinquent has higher loan amount which 
makes them harder to payoff.


## 3.1.3 Deliquency State and employment duration

```{r}
ggplot(data = loan_df, aes(x = DeliquencyState, y = EmploymentStatusDuration)) + 
      geom_boxplot()
```

The boxplot, in this case, does not make too much sense because of significant 
amount of data points that out of  quantiles from the mean. Because of that, 
I used the point plot to have visulize the employement duration.

## 3.2.1 Number of Investors and loan category
Because all listing categories are in numeric forms, it is necessary to map 
the english descriptvie name to them according to the data dictionary.

```{r}
#convert numeric values lvevel names to factor level names
#Input category names 
CategoryName <- c("Not Available", "Debt Consolidation", "Home Improvement", "Business", "Personal Loan", 
                     "Student Use", "Auto", "Other", "Baby&Adoption", "Boat", "Cosmetic Procedure", 
                      "Engagement Ring", "Green Loans","Household Expenses", "Large Purchases", "Medical/Dental", 
                     "Motorcycle", "RV", "Taxes", "Vacation", "Wedding Loans")
#create a new column and map full category name on this new column
loan_df$ListingCategory <- CategoryName[(loan_df$ListingCategory..numeric.)+1]

ggplot(data = loan_df, aes(ListingCategory, Investors)) + geom_boxplot() + coord_flip()
```

## 3.2.2 Number of Investors and credit score

Will higher credit score attract more investors? A scatter plot with trend line might be able answer this question.

```{r}
ggplot(data = loan_df, aes(x=CreditScoreRangeUpper, y=Investors)) + geom_point(shape=1) + 
        geom_smooth() + xlim(100, 1000)
```

Since it is barely possible for a person has credit score of 0, I only plotted 
loans with credit score above 0.
Although the relationship between credit score and is approximately position, 
it is noticeable that there is a downward trend after credit score go over 780.

## 3.2.3 Number of Investors and credit score

```{r}
ggplot(data = loan_df, aes(IncomeVerifiable, Investors)) + geom_boxplot()
```

Interestingly, weather the income verifiable does not have obvious influence 
on the number of investors. 

## 3.3.1 Loan original amount and loan category

```{r}
ggplot(data = loan_df, aes(ListingCategory, LoanOriginalAmount)) + geom_boxplot() + coord_flip()
```

## 3.3.2 Pearson Correlation

```{r}
ggcorr(data = loan_df, label = TRUE, size = 3, hjust = 1) 
```

From the correlation plot above, we found fairly strong correlations between 
Investors and loan amount, credit score and prosper score. These three variables 
will be explored deeper in next plot. However, it is also necessary to keep in 
mind that ProsperScore have high correlations of .4 with creditrange. 

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \ investigation. How did the feature(s) of interest vary with other features in \ the dataset?

1. For the deliquency of loans, the duration of employement and credit score seems have most obvious relationship with the deliquency state. Longer employement history and higher the credit score seems relevant to lesser deliquent loans.

2. As of the number of investors, the credit score and loan categories both have 
effects. The credit score has much more significant effects on the number 
compare to loan categories. The number of investors increased as the credit 
score goes up. 

### Did you observe any interesting relationships between the other features \ (not the main feature(s) of interest)?

The debt consolidation has highest median for loan amount amoung all other 
categories. Medical and dental loans does not own a high loan amount. 
The reason might be the residents of United States heavily rely on 
insurance and law service for health coverage. 

### What was the strongest relationship you found?

The strongest I found is the relationship between credit score and the 
number of investor. 


# 4. Multivariate Plots Section

## 4.1 Credit score, prosperscore, loan original amount and number of investors

```{r}
fit <- lm(Investors ~ LoanOriginalAmount + CreditScoreRangeUpper, data = loan_df) 
```

```{r}
```

```{r}
summary(fit)
```

The amateur regression model here does not represent the actual relationship 
between scores, loan amount and the number invester. Instead, p value less 
than 0.05 shows credit scores and loan amount do have statistically significant 
impact on the number of investers.
From the Q-Q plot, it can ben seen that the distribution of investors is not 
normal, which is also shwon in the univariate analysis. 
From the Leverage plot, there's a outlier located on the left of the graph. T
his group of points indicate that there are mulitple outliers while doing the regression.
From the residual plot, it can be seen that as the number of investors goes 
large, its become harder and harder to predict the number of investors. 

# 4.2 Deliquncy State vs loan amount and credit score

```{r}
ggplot(data = loan_df, aes(StatedMonthlyIncome, CreditScoreRangeUpper, colour = factor(DeliquencyState))) + 
geom_jitter(alpha = 0.1) + xlim(0, 10000)
```

# 4.3 Deliquncy State vs borrower's state and loan amount

```{r}
ggplot(loan_df, aes(BorrowerState, LoanOriginalAmount, fill = factor(DeliquencyState))) + geom_boxplot() + coord_flip() + theme(axis.text.y = element_text(size=5,vjust=1)) 
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Relationship between number of investors and credit scores, derived in 
bivariate analysis are used in conjuntion with the original loan amount in
multivariate. The result is significant, these two variables can be used 
together to built a regression model to predict the number of investors. 

### Were there any interesting or surprising interactions between features?

Seems the prosper score, which is the socre assigned by prosper loan 
marketplace has strong correlation with the credit score assigned by 
credit union. This is either because the prosper loan market place 
assign their score based on credit score from credit union or people's 
bevaior on the prosper market place reflect their general credit behavior.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

I tried to build a model based as: Investors ~ LoanOriginalAmount + 
CreditScoreRangeUpper. However, some spricification for linear regression 
is break, since the purpose is not to build a successful but to identify 
the general trend between variables, I didn't dived further onto this model. 

------

# 5. Final Plots and Summary

### Plot One

```{r echo=FALSE, Plot_One}
ggplot(data = loan_df, aes(StatedMonthlyIncome, CreditScoreRangeUpper, colour = factor(DeliquencyState))) + 
geom_jitter(alpha = 0.2) + xlim(0, 10000) + labs(title ="Deliquency state vs Stated Monthly Income and Credit Score", x = "Stated Monthly Income ($)")
```

Even the deliquency status of loans in significant impacted by the credit 
score of borrowers, the monthly income also plays a role in the behavior of 
deliquency status. For this plot, it is obvious that lower credit score 
implies higher chance of deliquent loans. Borrowers with higher monthly 
income will have higher credit score and higher chance of pay out their loans.

### Plot Two

```{r echo=FALSE, Plot_Two}
layout(matrix(c(1,2,3,4),2,2))
plot(fit, main="Investors vs Loan Amount + Credit Score")
```

From the Q-Q plot, it can ben seen that the distribution of investors is not
normal, which is also shwon in the univariate analysis. 
From the Leverage plot, there's a outlier located on the left of the graph. 
This group of points indicate that there are mulitple outliers while doing the regression.
From the residual plot, it can be seen that as the number of investors goes 
large, its become harder and harder to predict the number of investors. 

### Plot Three

```{r echo=FALSE, Plot_Three}
ggplot(data = loan_df, aes(x = DeliquencyState, y = LoanOriginalAmount)) + 
      geom_boxplot()+labs(title ="Deliquency state vs Stated Monthly Income and Credit Score", y = "Original Loan Amount ($)")
```

From this plot, we got a interesting finding that loans with higher loan 
original amount owns bettwe chance of good standing at the end. There couple 
theories for this behaviors. Firstly, this might because these loan has longer 
payment cycle so that they are still going through the pay back cycle. It coule 
also because borrowers borrowed a lot from loan companies tend to making better 
judgement of their financial situation.

------

# 6. Reflection

The Prosper Marketplace is the first P2P loan marketplace in the United States
and still operating today. This data set from Prosper Market place contains 81 
variables and 113,937 observations. I picked about 20 features for analyzing. 

Started from univariate analysis, I tried to understand the distribution of 
indivisual variables by using bar-plots and hitograms. I was expecting find 
much normal behaved variables but suprised by the log-normal and reversed 
lognormal behaviors found in these variables. For example, both the loan 
original amount and total credit line in past 7 years showing log normal 
behaviors. The number of investors ia also showing log-normal distribution 
in some sense. 

I was struggled while doing linear regresssion about the number of investors. 
Specificly, although the number of investors, credit score and original loan 
amount are positively correlated, all specifications for linera regression are 
failed to hold. However, the dianosis plot successfully dipicted issues with 
the regression and helped the analysis. The ANOVA table indicated that both 
the credit score and the original loan amount is significant when predicting 
the number of investors. 

Because of the good balance between categorical variables and numerical 
variables as well as the abundantness of information. I think this dataset 
would be a good resource for machine learning and neural network. I belive 
that after encode all categorical variables and apply machine learning 
algorithm, a impressive result is guaranteed.

## 7 Reference

https://yihui.name/knitr/options/

https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf

https://www.statmethods.net/stats/regression.html

https://stackoverflow.com/questions/41205489/different-fill-values-for-different-factors-in-ggplot2